---
title: "Methane in Hector"
author: "Natalie Crowe"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Building a Hector Core

Our first step is to create a core. Hector will use this to pull data from existing climate projections. Our projections will be from SSP245.

```{r  Hector-core, message=FALSE}
library(hector)
library(ggplot2)
library(dplyr)
ini_file <- system.file("input/hector_ssp245.ini", package = "hector")
core <- newcore(ini_file)
run(core)
```

## Methane Emissions Data and Graph

After creating and running the core we query it for methane-related outputs
between the years of 1745 and 2100. Hector has several options that relate to methane, including emissions, concentration, postindustrial, natural, soil lifetime and stratosphere lifetime.

```{r Emissions-data-frame}
results<- fetchvars(core, 1745:2100, c(EMISSIONS_CH4(),
                                       CONCENTRATIONS_CH4()))
invisible(shutdown(core))
head(results)
```

### Plot Results Methane Emissions

Now we have all the things we need to build a simple plot of methane emissions between the years 1745 and 2100. The `ggplot2` package is helpful for making graphs quickly and simply. We already loaded this package at the beginning of this file. 

```{r Emissions-plot}
results %>% 
  filter(variable == EMISSIONS_CH4()) %>% 
  ggplot(aes(x= year, y= value))+
    geom_line(color="blue")+
    labs(x = "Year",
         y = "Methane Emissions (Tg)",
         title = "Methane Emissions 1745-2100 in SSP245")
```

## Methane Concentrations

By changing the input `EMISSION_CH4()` to `CONCENTRATIONS_CH4()` we can replicate our code to get different visualized data.

```{r Concentration-graph}
results %>% 
  filter(variable == CONCENTRATIONS_CH4()) %>% 
  ggplot(aes(x= year, y= value))+
    geom_line(color= "red")+
    labs(x = "Year",
         y = "Methane Concentration (ppbv)",
         title = "Methane Concentration 1745-2100 in SSP245")
```

## Comparison Plots

We can put these two results into a single graph...
```{r Comparison-graph}
ggplot(data= results, aes(x= year, y= value, color= units))+
    geom_line()+
    labs(x = "Year",
         y = NULL,
         title = "Methane Concentration/Emission 1745-2100 in SSP245",
         color = "Units") + facet_wrap(~variable)
```
... to show our data side-by-side.
